---
title:  "Intro to R"
author: "Matthew Sisk <msisk1@nd.edu> & James Ng, PhD <james.ng@nd.edu>"
date:   "Navari Family Center for Digital Scholarship, University of Notre Dame"
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
  pdf_document: default
---

### You can use RStudio in the cloud! 
#### Sign up with google or github account at <a href="https://rstudio.cloud">rstudio.cloud</a>.

```{r}
library(tidyverse)
## you may need to install tidyverse first:
# install.packages('tidyverse')
rm(list=ls())
# setwd('/Users/jng2/Dropbox/Work/Library/CDS/R-RStudio/demo1hr')
```

### R has many baked-in example datasets. Let's load one of them. This is Fuel economy data from 1999 and 2008 for 38 popular models of car.
More info: https://ggplot2.tidyverse.org/reference/mpg.html

```{r}
datauto <- data.frame(mpg)
```

### Get file dimensions and store them in new objects.

```{r}
dim(datauto)
n.obs <- dim(datauto)[1]
n.cols <- dim(datauto)[2]
```

### Get a sense of the data.

```{r echo = T, results = 'hide'}
# get glimpse of data
str(datauto)
glimpse(datauto)
```

```{r echo = T, results = 'hide'}
# return first few rows
head(datauto)
```

```{r echo = T, results = 'hide'}
# get basic summary stats
summary(datauto)
```

### The variable class is categorical so let's factorize it. Summary() is now more useful when applied to a factor.
```{r}
datauto$f.class <- as.factor(datauto$class)
summary(datauto$f.class)
```

## Sort data by city mpg

```{r echo = T, results = 'hide'}
# ascending order
arrange(datauto, cty)
```

```{r echo = T, results = 'hide'}
# descendng order
arrange(datauto, desc(cty))
```

## Sort by manufacturer, model and year, and commit to memory.

```{r echo = T, results = 'hide'}
datauto <- arrange(datauto, manufacturer, model, year)
```

### Give me the names of all the manufacturers 

```{r}
unique( datauto[, 'manufacturer'] )
# equivalently: 
# unique( datauto$manufacturer )
```

### How many models per year did each manufacturer have? 

```{r echo = T, results = 'hide'}
datauto %>% 
  group_by(manufacturer, year) %>% 
  summarize( n.models = n_distinct(model) )
```

### Keep only the manufacturers with just one model.

```{r echo = T, results = 'hide'}
datauto %>% 
  group_by(manufacturer, year) %>% 
  filter( n_distinct(model) == 1 )
```

### Count number of vehicles by class per year and plot it

```{r}
classcounts <- datauto %>% group_by(f.class, year) %>% summarize(n = n())
ggplot(dat=classcounts, aes(y=n, x=f.class, fill=as.factor(year))) + 
  geom_bar(stat="identity")
```


### Compute mean cty, mean hwy and number of vehicles by class 

```{r}
datauto.means <- datauto %>% 
  group_by(f.class) %>% 
  summarize(mean.cty = mean(cty), mean.hwy = mean(hwy), n = n() )
datauto.means
```

### To produce this scatter plot.

```{r}
library(ggrepel)
plot1 <- ggplot(data=datauto.means, aes(x = mean.hwy, y = mean.cty)) +
  geom_point() +
  geom_text_repel(aes(label=f.class))
plot1
```

### Vary point color by mean displacement.

```{r}
plot1 <- plot1 + geom_point(aes(size=n))
plot1
```

### Find correlation between average combined cty and hwy mpg in 1999 vs 2008.
### To do this, need to 'spread' (reshape) the data.

```{r echo = T, results = 'hide'}
# first get average combined cty and hwy, call it mpg
datauto %>% mutate( mpg = (cty+hwy)/2 )
```

```{r echo = T, results = 'hide'}
# then summarize by manufacturer, model, year
tempdat <- datauto %>% mutate(mpg = (cty+hwy)/2) %>%
  group_by(manufacturer, model, year) %>%
  summarize( mpg=mean(mpg) )
tempdat
```

```{r echo = T, results = 'hide'}
# spread mpg into mpg_1999 and mpg_2008
tempdat %>% spread( key=year, value=mpg)
```

```{r echo = T, results = 'hide'}
# finally, rename 1999 and 2008 columns
dat.mpg.wide <- tempdat %>% spread( key=year, value=mpg) %>%
  rename( mpg_1999 = `1999`, mpg_2008 = `2008` )
dat.mpg.wide
```

### What's the relationship between fuel economy in 1999 and fuel economy in 2008?

```{r}
modl <- lm(formula = mpg_2008 ~ mpg_1999, data=dat.mpg.wide)
summary(modl)
```

### Want to learn more? Two great resources:
#### <a href="http://r4ds.had.co.nz/">R for Data Science</a> by Hadley Wickham
#### <a href="https://swcarpentry.github.io/r-novice-inflammation/">Programming with R</a> by Software Carpentry