---
title:  "Intro to R"
author: "Matthew Sisk <msisk1@nd.edu> & James Ng, PhD <james.ng@nd.edu>"
date:   "Navari Family Center for Digital Scholarship, University of Notre Dame"
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "G:\\My Drive\\Teaching\\Datasets")


```

## The Basics of using R

First, you should download and install R (https://cran.rstudio.com/) and RStudio (https://rstudio.com/products/rstudio/download/) if you have not done so yet. Both should work on any of the major opperating systems and be fairly easy to install

If it absolutly does not work, you can use RStudio in the cloud! 
Sign up with google or github account at <a href="https://rstudio.cloud">rstudio.cloud</a>.

Once the software is all set up, you should run RStudio. 

The first thing we will want to do is load the tidyverse package. This brings in a lot of the newer ways of manipulating data in R. all external libraries (called packages) are brought in with the *library()* command. If you have never installed it, you will need to run *install.packages("tidyverse")* first.

```{r tidyload}
library(tidyverse)
## you may need to install tidyverse first:
# install.packages('tidyverse')
```

Then we will want to clear out any existing data and set our working directory for any files we want to load or save. Note: your WD will be different and on windows computers you should use two backslashes between parts of the location
```{r wd}
rm(list=ls())
setwd("G:\\My Drive\\Teaching\\Datasets")
```



Then you are ready to load some data 

### Part 1: Loading Data

R has many baked-in example datasets. You can see a list of what is available in your version of R by running the command *data()* in your console


Let's load one of these datasets, mpg. This is Fuel economy data from 1999 and 2008 for 38 popular models of car.
More info: https://ggplot2.tidyverse.org/reference/mpg.html

```{r mpgLoad}
datauto <- mpg
```
Here, we have taken the built-in mpg dataset and made a copy of it into a local object called *datauto*. The arrow <- is how we assign values in R. An = also works in most cases. We can assign other variables the same way

```{r other variables}
twice <- 2
animal <- "goat"
moods <- c("happy", "sad", "indifferent")
```
The c() syntax is how R knows something is a list of values

Alternatively, you can load many other file types directly into R. The newer way of doing this is with the readr package (which comes as part of the tidyverse).
you can read comma-separated values (csv) files with read_csv
```{r readexternal}
census.data <- read_csv("State_ACS2016.csv")
```

Alternatively, you import excel files directly with read_exel, though you need to import the *readxl* package and specify the sheet.
```{r readxl}
library(readxl)
prisons <-  read_excel("US_Prisons.xls", sheet="US_Prisons")

```

In the Environment section of RStudio you can see the differences between values and the data we loaded. The datauto object is a data frame, which is the R version of a table. A datafame has a series of observations (rows) and variables (columns)

#2. Summarizing Data
There are a number of different ways of getting a sense of your data in R. Here are a few of them


To get the data frame dimensions and store them in new objects.

```{r}
dim(datauto)
n.obs <- dim(datauto)[1]
n.cols <- dim(datauto)[2]
```

To get a sense of the data use *glimpse()*.

```{r echo = T, results = 'hide'}
glimpse(datauto)
```
To return first few rows use *head()*

```{r echo = T, results = 'hide'}

head(datauto)
```
To get basic summary stats use *summary()*
```{r echo = T, results = 'hide'}

summary(datauto)
```

We can reference individual rows, colums and records within the data frame.
To reference an individual column use the name of the dataframe $ and the name of the column
```{r inside}
mean(datauto$cty)

```
You can also reference columns or rows by number or name within brackets. The row number/name comes first, followed by a comma, then the column number/name. If you are referencing 
```{r inside2}
print(datauto[4,"trans"])
datauto[4,]
max(datauto[,c("hwy","cty")])
```

##3. Manipulating Data

The command *unique()* will give all of the possible values

```{r}
unique( datauto[, 'manufacturer'] )
# equivalently: 
# unique( datauto$manufacturer )
```

Reassigning columns is done the same way variables are assigned.
Here, the variable *class* is categorical so let's factorize it. This basically makes it so there is not as much repetative text stored in the file and analyses run faster.
Summary() is now more useful when applied to a factor.
```{r}
datauto$f.class <- as.factor(datauto$class)
summary(datauto$f.class)
```

We can also sort and arrange our data off of any field. Here we sort by manufacturer, model and year.
```{r echo = T, results = 'hide'}
datauto <- arrange(datauto, manufacturer, model, year)
```



###ABANDON ALL HOPE
### Give me the names of all the manufacturers 

```{r}
unique( datauto[, 'manufacturer'] )
# equivalently: 
# unique( datauto$manufacturer )
```

### How many models per year did each manufacturer have? 

```{r echo = T, results = 'hide'}
datauto %>% 
  group_by(manufacturer, year) %>% 
  summarize( n.models = n_distinct(model) )
```

### Keep only the manufacturers with just one model.

```{r echo = T, results = 'hide'}
datauto %>% 
  group_by(manufacturer, year) %>% 
  filter( n_distinct(model) == 1 )
```

### Count number of vehicles by class per year and plot it

```{r}
classcounts <- datauto %>% group_by(f.class, year) %>% summarize(n = n())
ggplot(dat=classcounts, aes(y=n, x=f.class, fill=as.factor(year))) + 
  geom_bar(stat="identity")
```


### Compute mean cty, mean hwy and number of vehicles by class 

```{r}
datauto.means <- datauto %>% 
  group_by(f.class) %>% 
  summarize(mean.cty = mean(cty), mean.hwy = mean(hwy), n = n() )
datauto.means
```

### To produce this scatter plot.

```{r}
library(ggrepel)
plot1 <- ggplot(data=datauto.means, aes(x = mean.hwy, y = mean.cty)) +
  geom_point() +
  geom_text_repel(aes(label=f.class))
plot1
```

### Vary point color by mean displacement.

```{r}
plot1 <- plot1 + geom_point(aes(size=n))
plot1
```

### Find correlation between average combined cty and hwy mpg in 1999 vs 2008.
### To do this, need to 'spread' (reshape) the data.

```{r echo = T, results = 'hide'}
# first get average combined cty and hwy, call it mpg
datauto %>% mutate( mpg = (cty+hwy)/2 )
```

```{r echo = T, results = 'hide'}
# then summarize by manufacturer, model, year
tempdat <- datauto %>% mutate(mpg = (cty+hwy)/2) %>%
  group_by(manufacturer, model, year) %>%
  summarize( mpg=mean(mpg) )
tempdat
```

```{r echo = T, results = 'hide'}
# spread mpg into mpg_1999 and mpg_2008
tempdat %>% spread( key=year, value=mpg)
```

```{r echo = T, results = 'hide'}
# finally, rename 1999 and 2008 columns
dat.mpg.wide <- tempdat %>% spread( key=year, value=mpg) %>%
  rename( mpg_1999 = `1999`, mpg_2008 = `2008` )
dat.mpg.wide
```

### What's the relationship between fuel economy in 1999 and fuel economy in 2008?

```{r}
modl <- lm(formula = mpg_2008 ~ mpg_1999, data=dat.mpg.wide)
summary(modl)
```

### Want to learn more? Two great resources:
#### <a href="http://r4ds.had.co.nz/">R for Data Science</a> by Hadley Wickham
#### <a href="https://swcarpentry.github.io/r-novice-inflammation/">Programming with R</a> by Software Carpentry